---
title: "Hospital Analysis Report"
author: "Mark Ewing"
date: "October 10, 2016"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(plotly) #interactive graphics
library(DT) #interactive tables
library(choroplethr) #choropleth map

library(readxl)
library(dplyr)
library(tidyr)

hospitalData = read_excel("excel version.xlsx")
data(df_pop_state)

match_state = Vectorize(function(x){
  o = tolower(state.name[which(state.abb == x)])
  if(length(o) == 0) return(NA) else return(o)
})
```

# Inpatient Prospective Payment System

This analysis is based on data containing `r nrow(hospitalData)` rows of data covering `r length(unique(hospitalData[["DRG Definition"]]))` DRG Definitions.

## Where are the most discharges?

```{r, warning=F}
hospitalData %>% 
  group_by(`Provider State`) %>% 
  summarise(discharges = sum(` Total Discharges `)) %>% 
  mutate(region = unlist(match_state(`Provider State`))) %>% 
  left_join(df_pop_state,by=c("region")) %>% 
  mutate(value = discharges/value*1000) %>% 
  state_choropleth(title="Discharges per 1000 People by State")
```

The West Coast has significantly fewer discharges on a per capita basis while the Midwest and the South have higher rates.

### Provider discharges

```{r}
hospitalData %>% 
  mutate(totalMedicare = `Average Medicare Payments`*` Total Discharges `) %>% 
  group_by(`Provider Name`,`Provider State`) %>% 
  summarise(`Total Medicare Payments` = sum(totalMedicare),
    `Discharges` = sum(` Total Discharges `)) %>% 
  ungroup() %>% 
  plot_ly(y = `Total Medicare Payments`, x = Discharges, text = `Provider Name`, mode = 'markers')
```

We can see from this graph that the average Medicare payment of a discharge is about $10,000.  Some providers like Florida Hospital are lower while the New York Presbyterian Hospital are higher.

```{r}
hospitalData %>% 
  mutate(totalMedicare = `Average Medicare Payments`*` Total Discharges `) %>% 
  group_by(`Provider Name`,`Provider State`) %>% 
  summarise(`Total Medicare Payments` = sum(totalMedicare),
    `Discharges` = sum(` Total Discharges `)) %>% 
  ungroup() %>% 
  arrange(desc(Discharges)) %>% 
  datatable() %>% 
  formatCurrency('Total Medicare Payments') %>% 
  formatCurrency('Discharges','',digits=0) #formatting numbers with thousands commas
```

## Inpatient Codes

```{r}
hospitalData %>% 
  mutate(DRG = gsub("^[0-9]+ \\- | W/?O?.*CC$","",`DRG Definition`)) %>% 
  group_by(DRG) %>% 
  summarise(Discharges = sum(` Total Discharges `),
    `Maximum Payment` = max(` Average Total Payments `)) %>% 
  arrange(desc(`Discharges`)) %>% 
  datatable() %>% 
  formatCurrency('Maximum Payment') %>% 
  formatCurrency('Discharges','',digits=0) #formatting numbers with thousands commas
```

These codes have been collapsed to eliminate slight variations on treatments (e.g. with and without MCC/CC)

### Top 5 Codes

```{r}
hospitalData %>% 
  mutate(DRG = gsub("^[0-9]+ \\- | W/?O?.*CC$","",`DRG Definition`)) %>% 
  group_by(DRG) %>% 
  summarise(`Payment` = max(` Average Total Payments `),
    `Covered` = max(` Average Covered Charges `),
    `Medicare` = max(`Average Medicare Payments`)) %>% 
  arrange(desc(`Payment`)) %>% 
  slice(1:5) %>% 
  gather("Amount Type","Amount",-1) %>% 
  plot_ly(y = Amount, x = DRG, group = `Amount Type`, type='bar')

```

## Price / Volume

```{r}
hospitalData %>% 
  mutate(DRG = gsub("^[0-9]+ \\- | W/?O?.*CC$","",`DRG Definition`)) %>% 
  group_by(DRG) %>% 
  summarise(Discharges = sum(` Total Discharges `),
    `Maximum Payment` = max(` Average Total Payments `)) %>% 
  plot_ly(y = `Maximum Payment`, x = Discharges, text = DRG, mode='markers')
```

Most operations follow a typical price-volume curve - more frequent operations cost less due to their more routine nature.  Joint replacements are a bit of an exception (because of the physical cost of the replacement joint?) but otherwise there appear to be a few tiers of of operations, each with their own curve.